---
title: "Session 1 — Working with real social-science data in R"
subtitle: "R for Research in the Humanities and Social Sciences — CUSO"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: false
editor_options:
  chunk_output_type: inline
---

**Course:** R for Research in the Humanities and Social Sciences - CUSO 
**Format:** online, 2 hours  
**Dataset:** `carData::SLID` — Survey of Labour and Income Dynamics  

## What we will do

This first session is built around a real social-science dataset rather than a toy example. `SLID` contains observations from the 1994 wave of the Canadian Survey of Labour and Income Dynamics in Ontario. Each row is a respondent; the variables include hourly wages, years of education, age, sex, and language. There are also missing values, especially for wages.

That makes it useful for learning R because the structure resembles many research datasets in the social sciences, psychology, education, and related fields:

> **individuals → measurements → groups → missing data → summaries → figures**

The goal today is not to learn every feature of R. The goal is to understand what R is doing, line by line, and to acquire a small vocabulary that already lets us answer a real empirical question.

Our workflow will be:

> **Question → inspect → clean → transform → summarise → visualise → interpret**

The main empirical question will be:

> **How is education associated with hourly wages in this sample?**

We will also ask whether the pattern looks similar across groups. We will remain descriptive today: association is not causation.

---

# Before the session

You need **R** and **RStudio** installed.

Run this once before class:

```{r install-packages, eval=FALSE}
# install.packages() downloads and installs an R package.
# The package name is written between quotation marks because it is text.
#
# We need two packages for this session:
#   - tidyverse: tools for manipulating and visualising data;
#   - carData: the package that contains the SLID teaching dataset.
#
# We install them one at a time so that the syntax is easy to read.
# On your own computer, you normally need to install each package only once.

install.packages("tidyverse")
install.packages("carData")
```

During the course we will load the packages with `library()` or access a dataset directly from a package.

A useful distinction:

```{r}
# Installing and loading are two different operations.
#
# 1. install.packages("tidyverse") installs the package on the computer.
#    You usually do this only once.
#
# 2. library(tidyverse) loads the installed package into the current R session.
#    You do this again whenever you restart R and want to use the package.
#
# install.packages("tidyverse")   # normally NOT repeated every session
library(tidyverse)                # repeated in each new R session
```

Open the `R_Cuso_2026` folder as an RStudio project or working folder. We will avoid hard-coded paths such as:

```{r}
# Avoid hard-coded paths that only exist on your own computer.
# For example, this would make the script difficult to reuse elsewhere:
#
# setwd("C:/Users/myname/Desktop/project")
#
# Instead, open the whole project/folder in RStudio and use relative paths
# such as "data/my_file.csv" or "figures/my_plot.png".
```

Relative paths and projects make analyses easier to reproduce on another computer.

---

# 1. R is a calculator, but one that remembers things
We begin in the Console or in an R script.

Anything following `#` is a **comment**. R ignores it. Comments are for humans — including your future self.

```{r}
# R can evaluate arithmetic expressions directly.

# Addition
1 + 1

# Multiplication
3 * 4

# Division
10 / 4

# Powers: 2^3 means 2 × 2 × 2
2^3

# Parentheses control the order of operations.
# R calculates (2 + 3) first, then multiplies the result by 4.
(2 + 3) * 4
```

When you run one of these lines, R:

1. reads the expression;
2. evaluates it;
3. prints the result.

## Numbers and text are different

```{r}
# The value below is numeric: R can use it in arithmetic.
12

# Quotation marks turn the same characters into text.
# R now treats "12" as a character string, not as the number twelve.
"12"

# Numeric values can be added.
12 + 1
```

Now deliberately create an error:

```{r deliberate-error, error=TRUE}
# Run this line deliberately.
# It produces an error because "12" is text and 1 is numeric.
# The point is to see that an error message is information, not a disaster.
"12" + 1
```

Errors are normal. An error message tells us where R could not understand or execute an instruction.

## Store a value in an object

```{r}
# <- is the assignment operator.
# Read the next line as: "store the value 10 in an object called x".
x <- 10

# Typing the object's name asks R to display the value currently stored in it.
x

# Once an object exists, we can reuse it in another expression.
x + 5

# Assignment can also replace the previous value.
# After this line, x no longer contains 10: it contains 100.
x <- 100
x
```

Object names should tell us what they contain.

```{r}
# Short names are legal, but they quickly become hard to understand.
x <- 14

# Descriptive names make the meaning of the object visible in the code.
# Underscores are commonly used to separate words in R object names.
mean_years_education <- 14

# Display the value we just stored.
mean_years_education
```

## Functions

A function takes input, performs an operation, and returns output.

```{r}
# A function is called by writing its name followed by parentheses.

# sqrt() takes a number and returns its square root.
sqrt(16)

# round() can take several named arguments.
# Here x is the number to round and digits says how many decimals to keep.
round(x = 3.1415926, digits = 2)
```

The general shape is:

```text
function(argument_1, argument_2, ...)
```

You do not need to memorise every function. You need to learn how to read a function call.

```{r documentation, eval=FALSE}
# A question mark followed by a function name opens its help page in RStudio.
# Documentation is part of normal R work: you are not expected to memorise everything.
?round
```

---

# 2. Load the tools and the dataset
## Load the tidyverse

```{r}
# library() loads an installed package into the current R session.
#
# tidyverse is a collection of packages. In this session we mainly use:
#   - dplyr for data manipulation;
#   - ggplot2 for visualisation.
library(tidyverse)
```

You may see messages about packages being attached or functions being masked. That is normal.

## Load the SLID dataset

The dataset is distributed with the `carData` R package.

```{r}
# data() loads a dataset that is distributed inside an R package.
#
# "SLID" is the name of the dataset.
# package = "carData" tells R where to find it.
data("SLID", package = "carData")
```

We now have an object called `SLID`.

```{r}
# head() displays the first six rows by default.
# This is a quick way to check what the dataset actually looks like.
head(SLID)
```

For the rest of the session, we will use a lower-case object name. This does not change the data; it simply creates another reference to the same values.

```{r}
# as_tibble() converts the ordinary data frame into a tidyverse "tibble".
# The values do not change; a tibble simply prints a compact preview,
# which is much easier to read in the console or notebook.
slid <- as_tibble(SLID)

# Typing the object's name now displays a manageable preview.
slid
```

## What does one row mean?

This question should always come before statistical analysis.

```{r}
# nrow() asks: how many observations (rows) are there?
nrow(slid)

# ncol() asks: how many variables (columns) are there?
ncol(slid)

# dim() returns both dimensions at once:
# first the number of rows, then the number of columns.
dim(slid)
```

Here, **one row corresponds to one respondent** in this extract of the survey.

That means that when we calculate something such as a mean age, the observational unit is the respondent — not a province, a household, or a year.

---

# 3. Never analyse a dataset you have not inspected
## Look at names

```{r}
# names() returns the names of all columns in a data frame.
# Before analysing data, check that the variables you expect are actually present.
names(slid)
```

The variables are:

```text
wages       hourly wage rate
education   years of schooling
age         age in years
sex         recorded sex category
language    language category
```

## Look at the first rows

```{r}
# head(slid) shows the first six observations.
head(slid)

# A second argument lets us choose how many rows to display.
# Here we ask for the first ten.
head(slid, 10)
```

## Look at structure and data types

```{r}
# glimpse() gives a compact overview of the whole data frame.
# For each variable, look at:
#   - its name;
#   - its type (<dbl>, <fct>, ...);
#   - a few example values.
glimpse(slid)
```

You should see types such as:

```text
<dbl>  numeric value, usually allowing decimals
<int>  integer
<fct>  factor: a categorical variable with defined levels
<chr>  character/text
```

Data types matter. Taking the mean of `age` makes sense. Taking the mean of `language` does not.

Check individual variables:

```{r}
# The $ operator extracts one column from a data frame.
# class() tells us how R is storing that column.

class(slid$wages)     # numeric wage values
class(slid$age)       # numeric age values
class(slid$sex)       # categorical variable
class(slid$language)  # categorical variable
```

## A quick summary

```{r}
# summary() gives a quick variable-by-variable overview.
#
# For numeric variables it reports quantities such as the minimum,
# quartiles, median, mean and maximum.
# For categorical variables it reports counts by category.
# It also reports missing values when they are present.
summary(slid)
```

Do not rush past this output. Ask:

- Are values in plausible ranges?
- Are there missing observations?
- Which variables are continuous?
- Which are categorical?
- Does the coding match what we think the variable means?

This habit catches many errors before they become results.

---

# 4. Columns are vectors — and missing values are values too
A data-frame column can be extracted with `$`.

```{r}
# The $ operator extracts one column from a data frame.
# The result is a vector: one age value for each respondent.
#
# We use head(..., 10) so that we see only the first ten values,
# rather than printing thousands of ages.
head(slid$age, 10)
```

This is a **vector**: an ordered collection of values.

```{r}
# Store the age column in a separate object called age.
age <- slid$age

# head(..., 10) displays only the first ten values.
# This is easier to inspect than printing the entire vector.
head(age, 10)

# length() counts how many values the vector contains.
# Here that should match the number of rows in slid.
length(age)
```

We can apply functions to a vector.

```{r}
# Functions can summarise all values in a numeric vector.

# Arithmetic mean
mean(age)

# Median: half the observations are below it and half are above it.
median(age)

# Standard deviation: a common measure of dispersion around the mean.
sd(age)
```

Now try wages:

```{r}
# Try the same operation on wages.
# R returns NA rather than a number because at least one wage value is missing.
mean(slid$wages)
```

Why?

Because the wage variable contains missing values.

```{r}
# is.na() checks every value and returns TRUE when it is missing
# and FALSE when it is observed.
#
# Show only the first 20 results so that the output stays readable.
head(is.na(slid$wages), 20)

# In R, TRUE behaves like 1 and FALSE like 0 in a sum.
# Summing the logical vector therefore counts the missing wage values.
sum(is.na(slid$wages))

# Taking the mean of TRUE/FALSE values gives a proportion.
# This is the share of wage observations that are missing.
mean(is.na(slid$wages))
```

To compute a mean while excluding missing observations from **that calculation**:

```{r}
# na.rm = TRUE tells mean() to remove missing values for this calculation.
# It does NOT modify the dataset itself.
mean(slid$wages, na.rm = TRUE)
```

Read `na.rm = TRUE` as:

> "remove missing values before applying this function."

Other examples:

```{r}
# The same na.rm argument can be used with many summary functions.

# Median of the observed wages
median(slid$wages, na.rm = TRUE)

# Standard deviation of the observed wages
sd(slid$wages, na.rm = TRUE)

# Smallest observed wage
min(slid$wages, na.rm = TRUE)

# Largest observed wage
max(slid$wages, na.rm = TRUE)
```

Important research point: `na.rm = TRUE` is a computational instruction, **not a missing-data strategy**. In a real study we would also ask why wages are missing and whether missingness is related to other variables.

---

# 5. Manipulate observations with `dplyr`
Most research questions require us to keep, transform, or combine variables and observations.

The `dplyr` package, loaded with the tidyverse, uses verbs that correspond closely to those operations.

## `filter()` keeps rows

```{r}
# filter() keeps ROWS that satisfy a condition.
#
# The first argument is the data frame.
# age >= 30 is the condition: keep a row when it is TRUE.
filter(slid, age >= 30)
```

A condition produces TRUE/FALSE values.

```{r}
# Before using filter(), it helps to see what the condition itself produces.
# R evaluates age >= 30 once for every respondent.
# The result is a vector of TRUE and FALSE values.
#
# Again, show only the first 20 values to keep the output readable.
head(slid$age >= 30, 20)
```

`filter()` retains the rows for which the condition is TRUE.

Several conditions can be combined.

```{r}
# Several conditions separated by commas inside filter() are combined with AND.
# A row is kept only if ALL three conditions are true.

filter(
  slid,             # start from the slid data frame
  age >= 30,        # age is at least 30
  age <= 40,        # age is at most 40
  !is.na(wages)     # ! means NOT: wage is not missing
)
```

Useful operators:

```text
==   equal to
!=   not equal to
>    greater than
<    less than
>=   greater than or equal to
<=   less than or equal to
!    logical NOT
&    AND
|    OR
```

For a categorical variable:

```{r}
# For categorical variables we usually test equality with ==.
# This keeps only rows whose language category is exactly "French".
#
# Be careful: = and == have different roles in R.
filter(slid, language == "French")
```

## `select()` keeps columns

```{r}
# select() keeps COLUMNS rather than rows.
# Here we create a temporary view containing only four variables.
select(
  slid,
  wages,
  education,
  age,
  language
)
```

## `arrange()` sorts rows

```{r}
# arrange() changes the order of the rows.

# Sort from the smallest wage to the largest.
arrange(slid, wages)

# desc() reverses the order:
# sort from the largest wage to the smallest.
arrange(slid, desc(wages))
```

Notice that none of these commands modifies `slid`. They return a new result.

---

# Short break
Take a short break away from the screen.

---

# 6. The pipe: write code in the order you think
Suppose we want to:

1. start with `slid`;
2. keep respondents with an observed wage;
3. keep only a few variables;
4. sort from highest to lowest wage.

We could nest functions, but it quickly becomes difficult to read.

The native R pipe `|>` means approximately **"then"**.

```{r}
# The native R pipe |> passes the result of one step into the next step.
# Read it aloud as "then".

slid |>                              # start with the full dataset
  filter(!is.na(wages)) |>           # THEN keep rows with an observed wage
  select(wages, education, age, language) |>  # THEN keep four columns
  arrange(desc(wages))               # THEN sort wages from high to low
```

Read the code aloud:

> Take `slid`, then filter, then select, then arrange.

The following two pieces of code are equivalent:

```{r}
# Without a pipe, the data frame is written explicitly as the first argument.
filter(slid, age >= 30)
```

```{r}
# With a pipe, slid is passed automatically into filter().
# This produces exactly the same result as the previous command.
slid |>
  filter(age >= 30)
```

The pipe becomes useful when the analysis has several consecutive steps.

## Save the result

```{r}
# A pipeline returns a result just like any other R expression.
# Using <- lets us save that result in a new object.

slid_complete_wage <- slid |>
  filter(!is.na(wages))

# The original dataset has not changed.
# Compare its number of rows with the new filtered dataset.
nrow(slid)
nrow(slid_complete_wage)
```

We have **not** overwritten the original data. Keeping the raw object unchanged is usually a good habit.

## `mutate()` creates variables

A common task in research is deriving a variable from existing measurements.

For example, create broad age groups:

```{r}
# mutate() adds or transforms columns while keeping the rows.
#
# We create a new variable called age_group from the numeric age variable.

slid_with_age_group <- slid |>
  mutate(
    age_group = case_when(
      # case_when() tests conditions from top to bottom.
      # The first matching condition determines the label.
      age < 30 ~ "under 30",
      age < 45 ~ "30–44",
      age < 60 ~ "45–59",
      age >= 60 ~ "60+"
    )
  )

# Inspect the result by showing only age and the newly created age_group.
# head(20) limits the output to the first 20 rows.
slid_with_age_group |>
  select(age, age_group) |>
  head(20)
```

The original `slid` object still has only the original variables:

```{r}
# names(slid) shows that the original object is unchanged:
# it still contains only the original variables.
names(slid)
```

The modified object contains the new one:

```{r}
# The new object contains one additional variable: age_group.
names(slid_with_age_group)
```

### Mini exercise

Create an object called `working_sample` that:

- keeps only respondents with observed wages;
- keeps only `wages`, `education`, `age`, `sex`, and `language`;
- sorts rows from highest to lowest wage.

Try it before opening the solution.

<details>
<summary>Solution</summary>

```{r}
# Start from the original data.
working_sample <- slid |>
  # Keep only respondents for whom wage is observed.
  filter(!is.na(wages)) |>
  # Keep the five variables relevant to the exercise.
  select(wages, education, age, sex, language) |>
  # Sort the resulting rows from highest to lowest wage.
  arrange(desc(wages))

# Display the object we have just created.
working_sample
```

</details>

---

# 7. Go from individual observations to evidence about groups
## Count observations

```{r}
# count(variable) counts how many rows belong to each category.
#
# The output has:
#   - one row per language category;
#   - a column called n containing the count.
slid |>
  count(language)
```

`count()` is useful whenever you need to understand the composition of a dataset.

We can immediately turn counts into percentages of **this sample**:

```{r}
# First count the respondents in each language category.
slid |>
  count(language) |>
  # count() created a column called n.
  # sum(n) is the total sample size.
  # Dividing each group count by the total gives its proportion;
  # multiplying by 100 expresses that proportion as a percentage.
  mutate(
    percent = 100 * n / sum(n)
  )
```

Be precise in interpretation: these are proportions in the dataset we are analysing. Survey weights and sampling design would matter before treating them as population estimates.

## One summary for the entire dataset

```{r}
# summarise() reduces many rows to a smaller set of summary statistics.
# Without group_by(), the calculations use the whole dataset.

slid |>
  summarise(
    # Mean age across all respondents with an observed age
    mean_age = mean(age, na.rm = TRUE),

    # Mean wage across respondents with an observed wage
    mean_wage = mean(wages, na.rm = TRUE)
  )
```

`summarise()` reduces many rows to one or more summary values.

## Summaries by group

Now ask:

> What do observed wages look like in each language category?

```{r}
# group_by(language) tells R to treat each language category separately.
# summarise() is then calculated once inside each group.

slid |>
  group_by(language) |>
  summarise(
    # n() counts all rows in the current group.
    n = n(),

    # Count only respondents whose wage is not missing.
    n_wage = sum(!is.na(wages)),

    # Mean wage among observed wage values.
    mean_wage = mean(wages, na.rm = TRUE),

    # Median wage among observed wage values.
    # The median is often useful when a distribution is skewed.
    median_wage = median(wages, na.rm = TRUE),

    # Standard deviation describes dispersion around the mean.
    sd_wage = sd(wages, na.rm = TRUE)
  )
```

The logic is important:

```text
group_by(language)
        ↓
English observations  → summarise()
French observations   → summarise()
Other observations    → summarise()
        ↓
one row per group
```

Now look at education:

```{r}
# The same group_by() + summarise() logic can be reused
# with a different outcome variable.

slid |>
  group_by(language) |>
  summarise(
    # Number of respondents in the language group
    n = n(),

    # Average years of education in that group
    mean_education = mean(education, na.rm = TRUE),

    # Dispersion of years of education in that group
    sd_education = sd(education, na.rm = TRUE)
  )
```

A crucial lesson: a difference between group means is a **description**. It does not by itself tell us why the groups differ.

---

# 8. Make the observations visible with `ggplot2`
`ggplot2` builds figures from layers. Instead of memorising complete plotting commands, we will construct a figure step by step.

The three pieces to remember are:

```text
data        which data frame?
aesthetics  which variables become x, y, colour, etc.?
geometry    points, bars, lines, boxes, ...?
```

## Step 1 — declare the data

```{r}
# ggplot() starts a plot.
# At this stage we give it only the dataset.
# Because we have not yet defined axes or marks, the plot is empty.
ggplot(data = slid)
```

Nothing is visible yet because we have not said what to plot.

## Step 2 — map a variable to an axis

```{r}
# aes() defines an aesthetic mapping:
# it tells ggplot which data variable should control a visual property.
#
# Here the wages variable is mapped to the x-axis.
# We still have not chosen how the observations should be drawn.
ggplot(
  data = slid,
  aes(x = wages)
)
```

Still no marks: we have defined an axis but not a geometry.

## Step 3 — add a geometry

```{r}
# geom_histogram() chooses the geometry:
# it divides a numeric variable into bins and counts observations in each bin.
#
# ggplot(...) defines the data and mapping.
# The + sign adds another layer to the plot.
ggplot(data = slid, aes(x = wages)) +
  geom_histogram()
```

You may see a warning about removed rows containing missing values. That warning is useful: `ggplot2` is telling us that some wage observations could not be drawn.

We can make the analytic choice explicit:

```{r}
# We can remove missing wage values explicitly before plotting.
# This makes the data-selection step visible in the code.

slid |>
  filter(!is.na(wages)) |>       # keep only observed wages
  ggplot(aes(x = wages)) +       # map wage to the x-axis
  geom_histogram()               # draw a histogram
```

## Relationship between two numerical variables

Our main question concerns education and wages.

```{r}
# A scatterplot represents the relationship between two numeric variables.
#
# education is mapped to x, wages to y.
# Each point corresponds to one respondent with usable values.
ggplot(
  slid,
  aes(x = education, y = wages)
) +
  geom_point()
```

With thousands of observations, points overlap. We can make them partly transparent.

```{r}
# With many respondents, points overlap.
# alpha controls transparency:
#   1    = fully opaque
#   0.25 = mostly transparent
#
# Because alpha is a fixed visual setting rather than a data variable,
# it is written outside aes().
ggplot(slid, aes(x = education, y = wages)) +
  geom_point(alpha = 0.25)
```

`alpha = 0.25` is **not** mapped to a variable. It is a fixed visual setting, so it is written outside `aes()`.

Compare:

```{r}
# FIXED visual property:
# every point receives exactly the same transparency.
#
# alpha is outside aes() because it is not linked to a variable.
ggplot(slid, aes(x = education, y = wages)) +
  geom_point(alpha = 0.25)
```

with:

```{r}
# MAPPED visual property:
# colour is inside aes() because colour now represents the variable language.
#
# ggplot automatically creates a legend for mapped aesthetics.
ggplot(
  slid,
  aes(
    x = education,
    y = wages,
    colour = language
  )
) +
  geom_point(alpha = 0.25)
```

This distinction — **inside vs outside `aes()`** — is one of the most important things to understand in `ggplot2`.

## Add a descriptive trend

```{r}
# Start with the scatterplot.
ggplot(slid, aes(x = education, y = wages)) +
  # Draw the individual respondents with transparent points.
  geom_point(alpha = 0.20) +
  # Add a straight fitted line.
  # method = "lm" asks for a linear model.
  # We use it here only as a descriptive summary of the association;
  # the statistical model itself will be studied in Session 2.
  geom_smooth(method = "lm")
```

At this stage the line is a visual summary of association. Do not interpret it as evidence that education itself *causes* a particular wage increase.

## Compare groups without putting everything on one plot

Facets create one panel per category.

```{r}
# facet_wrap() makes separate panels for categories of a variable.
# The same x and y mapping is reused in every panel.

ggplot(slid, aes(x = education, y = wages)) +
  geom_point(alpha = 0.20) +
  geom_smooth(method = "lm") +
  # ~ language means: create one panel for each language category.
  facet_wrap(~ language)
```

The formula:

```{r}
# In formulas used by functions such as facet_wrap(),
# the tilde ~ can be read roughly as "according to".
#
# Here: make panels according to language.
~ language
```

means: create panels according to values of `language`.

## Add human-readable labels

```{r}
# A plot can be built step by step and stored in an object.
wage_plot <- slid |>
  # Be explicit that only observed wages are used in the plot.
  filter(!is.na(wages)) |>
  # Define the variables on the x and y axes.
  ggplot(aes(x = education, y = wages)) +
  # Draw individual respondents.
  geom_point(alpha = 0.20) +
  # Add a descriptive linear trend.
  geom_smooth(method = "lm") +
  # Put language categories in separate panels.
  facet_wrap(~ language) +
  # labs() replaces technical variable names with labels for readers.
  labs(
    title = "Education and hourly wages in the SLID sample",
    subtitle = "Ontario, 1994; panels show recorded language category",
    x = "Years of education",
    y = "Hourly wage",
    caption = "Source: Survey of Labour and Income Dynamics (SLID)"
  ) +
  # theme_minimal() changes the overall visual appearance.
  theme_minimal()

# Display the finished plot.
wage_plot
```

Notice that a plot can itself be stored in an object:

```{r}
# Plots are R objects too.
# class() shows the class that R uses for the stored plot object.
class(wage_plot)
```

## Save a publication-quality image

Research code should be able to reproduce output files as well as results on screen.

```{r}
# dir.create() creates a folder.
# showWarnings = FALSE prevents a warning if the folder already exists.
dir.create("figures", showWarnings = FALSE)

# ggsave() writes a ggplot object to a file.
# This is reproducible: running the code again recreates the same output.
ggsave(
  filename = "figures/education_and_wages.png",  # output path
  plot = wage_plot,                              # plot object to save
  width = 8,                                     # width in inches
  height = 5,                                    # height in inches
  dpi = 300                                      # image resolution
)
```

This is much more reproducible than manually exporting the figure from RStudio.

---

# 9. Final challenge
Choose one question. If time is short, begin it now and finish after class.

### A — Education and wages

> How does hourly wage vary with years of education in this sample?

### B — Age and wages

> What does the relationship between age and observed wages look like?

### C — Group composition

> How are respondents distributed across language categories, and how do their observed wage distributions differ?

For your chosen question, produce:

1. one filtered or transformed dataset if necessary;
2. one summary table;
3. one figure;
4. one sentence describing what the data show.

A possible skeleton:

```{r}
# STEP 1 — choose the observations needed for the analysis.
analysis_data <- slid |>
  filter(!is.na(wages))

# STEP 2 — turn individual observations into a summary table.
summary_table <- analysis_data |>
  group_by(language) |>
  summarise(
    n = n(),
    mean_wage = mean(wages),
    median_wage = median(wages)
  )

# Display the summary table.
summary_table

# STEP 3 — make the relationship visible.
analysis_plot <- analysis_data |>
  ggplot(aes(x = education, y = wages)) +
  geom_point(alpha = 0.20) +
  geom_smooth(method = "lm") +
  theme_minimal()

# Display the plot.
analysis_plot
```

A careful descriptive sentence would look like:

> "In this sample, respondents with more years of education tend to report higher observed hourly wages."

That is different from:

> "Additional education causes higher wages."

The second statement requires a causal research design that we have not established.

---

# 10. What you now know

The amount of syntax introduced today is deliberately small.

```text
# Inspect
head()
glimpse()
summary()
nrow()
names()

# Missingness
is.na()

# Manipulate
filter()
select()
arrange()
mutate()

# Group and summarise
count()
group_by()
summarise()

# Visualise
ggplot()
geom_histogram()
geom_point()
geom_smooth()
facet_wrap()
labs()

# Save output
ggsave()
```

The more important idea is the workflow:

```text
What is one row?
      ↓
What are the variables and their types?
      ↓
What is missing?
      ↓
Which observations do I need?
      ↓
What numerical summary answers my question?
      ↓
What figure lets me see the observations?
      ↓
What can I legitimately conclude?
```

In Session 2 we will move from a visible association to **statistical evidence**: estimates, uncertainty, confidence intervals, and regression models.

---

# Optional: how this translates to your own CSV file

Today we used a dataset bundled with an R package so that everybody could run exactly the same code. Your own research data will often arrive as CSV or Excel files.

For a CSV file, the workflow changes only at the import step:

```{r import-examples, eval=FALSE}
# read_csv() imports a comma-separated text file as a data frame.
# The path is relative to the project folder.
#
# This line is an example: it will work only if that file actually exists.
my_data <- read_csv("data/my_data.csv")

# Always inspect imported data before analysing it.
glimpse(my_data)
summary(my_data)
```

For an Excel file:

```{r excel-example, eval=FALSE}
# Excel support comes from the readxl package.
# Install it once on your computer if necessary:
# install.packages("readxl")

# Load the package into the current R session.
library(readxl)

# read_excel() imports an Excel workbook.
# Again, this is an example path: replace it with the path to your own file.
my_data <- read_excel("data/my_data.xlsx")
```

Everything that follows — `filter()`, `mutate()`, `group_by()`, `ggplot()` — works with the same logic.





## Dataset source

`SLID` is distributed in the `carData` package and is described as data from the 1994 wave of the Canadian Survey of Labour and Income Dynamics for Ontario, prepared from the public-use dataset made available by Statistics Canada.
