---
title: "Session 2 — From description to statistical evidence"
output:
  html_document:
    toc: true
    toc_depth: 2
editor_options:
  chunk_output_type: inline
---

**Course:** R for Research  
**Format:** online, 2 hours  
**Language:** English  
**Dataset:** `carData::SLID` — Survey of Labour and Income Dynamics  

## What we will do

In Session 1, we learned how to inspect, transform, summarise and visualise a real social-science dataset.

Today we move one step further:

> **How do we go from a pattern in a sample to an estimated relationship with uncertainty?**

We will continue to use the 1994 Ontario extract from the Canadian Survey of Labour and Income Dynamics (`SLID`). The main empirical question will be:

> **How is education associated with hourly wages, and how does that association change when we also consider age and other respondent characteristics?**

This is a useful social-science example because it lets us move naturally from descriptive statistics to confidence intervals, group comparisons, simple regression and multiple regression.

The goal is not to turn statistical output into a ritual. At every step we will ask:

> **What quantity did we estimate, in what units, and what can we reasonably say about it?**

A final caution before we start: these data come from **Ontario in 1994**. We will use them to learn statistical reasoning. We should not treat the results as estimates of present-day Canadian society, and we are not using the survey design or weights required for formal population inference.

---

# Learning objectives

By the end of the session, you should be able to:

- distinguish **variation in the data** from **uncertainty around an estimate**;
- calculate and interpret a standard error and a confidence interval;
- compare two groups using differences and uncertainty, not only a p-value;
- understand a linear regression formula of the form `outcome ~ predictor`;
- fit a model with `lm()`;
- interpret a regression coefficient in the units of the data;
- understand how R represents categorical predictors;
- fit a multiple regression model with several predictors;
- explain what “holding other variables constant” means;
- recognise that statistical adjustment does not by itself establish causality;
- write a short result sentence that matches the evidence actually produced.

---

# Before the session

We use the same packages as in Session 1.

If the packages are not installed on your computer yet, install them once:

```{r, eval=FALSE}
# install.packages() downloads and installs a package from CRAN.
# The package name is text, so it is written between quotation marks.
#
# Installation is normally a ONE-TIME operation on a computer.
# These lines do not need to be run every time you start R.

install.packages("tidyverse")
install.packages("carData")
```

At the beginning of a new R session, load the tools and data we need:

```{r}
# library() does NOT install a package.
# It loads an already-installed package into the current R session.
# This is why library(tidyverse) is normally run again after restarting R.

library(tidyverse)

# data() loads a dataset that is distributed inside an installed package.
# Here the dataset is called SLID and it lives in the carData package.
data("SLID", package = "carData")

# Create a lower-case object name that we will use throughout the course.
# This leaves the original SLID object unchanged.
slid <- SLID
```

Check that everything worked:

```{r}
# glimpse() gives a compact view of:
# - the number of rows and columns;
# - variable names;
# - variable types;
# - a few example values.
glimpse(slid)

# summary() adapts its output to each variable type.
# For numeric variables it shows quantities such as the median and quartiles.
# For factors it shows counts by category.
summary(slid)
```

---

# 1. Recap: describe before you model

Our main outcome today is hourly wage.

Before fitting any model, inspect the variable.

```{r}
# is.na() produces TRUE for each missing wage and FALSE otherwise.
# sum() then counts the TRUE values, because TRUE behaves like 1 here.
sum(is.na(slid$wages))

# Taking the mean of TRUE/FALSE values gives the proportion that is TRUE.
# This therefore gives the proportion of wage observations that are missing.
mean(is.na(slid$wages))

# summary() describes the observed values and also reports missingness.
summary(slid$wages)
```

Remember that `NA` is not zero. It means that the value is unavailable.

Let us make the distribution visible.

```{r}
# ggplot() starts the figure and tells R which dataset to use.
# aes(x = wages) maps the wages variable to the horizontal axis.
ggplot(slid, aes(x = wages)) +
  # geom_histogram() divides wages into intervals and counts observations.
  # binwidth = 2 means each bar covers a two-unit wage interval.
  # boundary = 0 makes the bin boundaries start at multiples of 2 from zero.
  geom_histogram(binwidth = 2, boundary = 0) +
  # labs() gives the plot labels that a reader will actually see.
  labs(
    x = "Hourly wage",
    y = "Number of respondents",
    title = "Distribution of observed hourly wages"
  ) +
  # theme_minimal() changes presentation, not the underlying data.
  theme_minimal()
```

The distribution is not perfectly symmetric. That immediately tells us something important:

> A mean is useful, but it does not describe the whole distribution.

Compare mean and median:

```{r}
slid |>
  # summarise() reduces many rows to a small set of summary statistics.
  summarise(
    # na.rm = TRUE tells mean() to ignore missing wage values for this calculation.
    mean_wage = mean(wages, na.rm = TRUE),

    # The median is the middle observed wage after ordering the values.
    median_wage = median(wages, na.rm = TRUE),

    # The standard deviation describes how dispersed individual wages are
    # around their mean.
    sd_wage = sd(wages, na.rm = TRUE)
  )
```

## First principle of Session 2

A statistical model does not replace descriptive analysis.

A useful workflow is:

```text
inspect → visualise → estimate → quantify uncertainty → interpret
```

---

# 2. From a sample mean to uncertainty

Suppose we want to estimate the mean hourly wage in the observations for which wage is available.

First create a clean wage vector:

```{r}
# slid$wages extracts the wages column as a vector.
# !is.na(slid$wages) is TRUE only where the wage is observed.
# The square brackets keep only those observed positions.
wage_vector <- slid$wages[!is.na(slid$wages)]

# length() tells us how many observed wage values remain.
length(wage_vector)
```

Now calculate the sample mean and standard deviation.

```{r}
# The mean describes the centre of the observed wage values.
mean(wage_vector)

# The standard deviation describes variation between individual respondents.
sd(wage_vector)
```

These two quantities answer different questions.

- **Mean:** where is the centre of the observed values?
- **Standard deviation (SD):** how dispersed are individual observed values around that mean?

But if we are interested in the mean as an **estimate**, we also want to know how precisely it has been estimated.

## Standard error

For a simple mean, the standard error is:

```text
SE = SD / sqrt(n)
```

Let us calculate every step explicitly.

```{r}
# Number of non-missing wage observations.
n_wage <- length(wage_vector)

# Estimated mean wage.
mean_wage <- mean(wage_vector)

# Variation between individual observed wages.
sd_wage <- sd(wage_vector)

# sqrt(n_wage) is the square root of the sample size.
# Dividing the SD by sqrt(n) gives the standard error of the mean.
se_wage <- sd_wage / sqrt(n_wage)

# Typing an object's name asks R to print its current value.
n_wage
mean_wage
sd_wage
se_wage
```

The distinction is fundamental:

```text
SD  = variability among observations
SE  = uncertainty around the estimated mean
```

A dataset can contain highly variable people and still estimate its mean quite precisely if the sample is large.

## Approximate 95% confidence interval

With a large sample, a common approximation is:

```text
estimate ± 1.96 × standard error
```

```{r}
# Move 1.96 standard errors below the estimated mean.
lower_ci <- mean_wage - 1.96 * se_wage

# Move 1.96 standard errors above the estimated mean.
upper_ci <- mean_wage + 1.96 * se_wage

lower_ci
upper_ci
```

We can do all of this in one tidyverse pipeline:

```{r}
wage_summary <- slid |>
  summarise(
    # Count only rows where wages are observed.
    n = sum(!is.na(wages)),

    # Estimate the mean wage among those observed values.
    mean_wage = mean(wages, na.rm = TRUE),

    # Describe variation between respondents.
    sd_wage = sd(wages, na.rm = TRUE),

    # Use statistics created earlier in the SAME summarise() call.
    # Here n and sd_wage already exist when se_wage is calculated.
    se_wage = sd_wage / sqrt(n),

    # Construct the approximate 95% confidence interval.
    lower = mean_wage - 1.96 * se_wage,
    upper = mean_wage + 1.96 * se_wage
  )

# Print the resulting one-row summary table.
wage_summary
```

### What a confidence interval is not

A 95% confidence interval around the mean is **not**:

> “the interval containing 95% of respondents.”

It is an interval expressing uncertainty around an **estimated quantity** — here, the mean.

---

# 3. Comparing groups: estimate the difference

A very common social-science question is whether two groups differ on an outcome.

The SLID data contain a variable called `sex`, with categories recorded in the dataset.

First inspect it.

```{r}
# levels() shows the possible categories of a factor and their order.
# The order matters later because regression uses one level as a reference.
levels(slid$sex)

# count() returns one row per category with the number of observations.
slid |>
  count(sex)
```

Now visualise wages by recorded sex category.

```{r}
ggplot(slid, aes(x = sex, y = wages)) +
  # A boxplot displays the median, quartiles and potential outliers.
  # outlier.alpha changes only how transparent the outlier points appear.
  geom_boxplot(outlier.alpha = 0.25) +
  labs(
    x = "Recorded sex category",
    y = "Hourly wage",
    title = "Observed hourly wages by recorded sex category"
  ) +
  theme_minimal()
```

Do not jump straight to a test. First calculate quantities we can interpret directly.

```{r}
sex_summary <- slid |>
  # group_by() means that summarise() will now work separately for each sex category.
  group_by(sex) |>
  summarise(
    # Count only respondents whose wage is observed.
    n = sum(!is.na(wages)),

    # Mean observed wage in the group.
    mean_wage = mean(wages, na.rm = TRUE),

    # Median observed wage in the group.
    median_wage = median(wages, na.rm = TRUE),

    # Variation between observed wages inside the group.
    sd_wage = sd(wages, na.rm = TRUE),

    # Precision of the estimated group mean.
    se_wage = sd_wage / sqrt(n)
  )

sex_summary
```

This gives us two estimated means. We can now ask about their difference.

## A two-sample t-test

R can estimate the difference and its uncertainty with `t.test()`.

```{r}
# Formula syntax in R usually puts:
#   outcome ~ explanatory variable
#
# Here R compares wages across the categories of sex.
wage_ttest <- t.test(
  wages ~ sex,
  data = slid
)

# Print the complete test output.
wage_ttest
```

The output contains several quantities:

- estimated group means;
- a confidence interval for their difference;
- a test statistic;
- degrees of freedom;
- a p-value.

For research interpretation, a useful order is:

> **difference → uncertainty → substantive meaning → p-value if useful**

rather than:

> **p < .05 → stop thinking**

## What a p-value does not tell you

A p-value is not:

- the probability that the null hypothesis is true;
- the probability that the result is “due to chance”;
- the size of an effect;
- the practical importance of a difference.

With large samples, even small differences can produce small p-values.

### Important substantive caution

If two groups have different observed mean wages, this descriptive comparison does **not** tell us why.

For example, the groups may also differ in age, education, occupation, work experience, hours, sector, or many variables that are absent from this small teaching dataset.

That question motivates regression.

---

# Short break

---

# 4. Regression: one language for many research questions

Regression gives us a general language for estimating relationships between variables.

We will first create one analysis dataset and use the **same observations** in all models we want to compare.

This is an important research habit: otherwise a coefficient can change simply because the set of observations changed.

```{r}
analysis_data <- slid |>
  # Keep only variables that will appear in today's models.
  select(wages, education, age, sex, language) |>

  # drop_na() removes a row if ANY of these retained variables is missing.
  # Using one common analysis sample makes later model comparisons cleaner.
  drop_na()

# nrow() checks how many respondents remain after removing incomplete rows.
nrow(analysis_data)

# Inspect the variables and types in the resulting analysis dataset.
glimpse(analysis_data)
```

## Our first regression question

> How is education associated with hourly wages?

Before modelling, visualise the data.

```{r}
ggplot(
  analysis_data,
  aes(x = education, y = wages)
) +
  # Education is measured in whole years, so points would otherwise sit
  # directly on top of one another.
  # width adds a tiny horizontal displacement purely for visibility.
  geom_jitter(width = 0.12, height = 0, alpha = 0.10) +
  labs(
    x = "Years of education",
    y = "Hourly wage",
    title = "Education and observed hourly wages"
  ) +
  theme_minimal()
```

Add a linear trend.

```{r}
ggplot(
  analysis_data,
  aes(x = education, y = wages)
) +
  geom_jitter(width = 0.12, height = 0, alpha = 0.10) +

  # method = "lm" asks ggplot2 to fit a linear model.
  # The line is therefore the visual counterpart of the model fitted below.
  geom_smooth(method = "lm") +

  labs(
    x = "Years of education",
    y = "Hourly wage",
    title = "Education and observed hourly wages"
  ) +
  theme_minimal()
```

The shaded band shows uncertainty around the estimated **mean relationship**.

## Fit the model with `lm()`

```{r}
# lm() means "linear model".
# The formula wages ~ education reads:
# "model wages as a function of education".
model_1 <- lm(
  wages ~ education,
  data = analysis_data
)
```

Nothing is printed because we assigned the model to an object called `model_1`.

Ask R to summarise it:

```{r}
# summary() displays several diagnostics and a coefficients table.
# For now, focus mainly on Estimate, Std. Error and the coefficient names.
summary(model_1)
```

The output is long. For now, focus on the **coefficients** table.

We can also request only the estimated coefficients:

```{r}
# coef() extracts the estimated intercept and slope from the fitted model.
coef(model_1)
```

And their confidence intervals:

```{r}
# confint() gives a confidence interval for each model coefficient.
confint(model_1)
```

## Read the model in plain English

The model can be written conceptually as:

```text
expected wage = intercept + slope × education
```

The coefficient for `education` is the **slope**.

Suppose, only as an illustration, that the coefficient were `0.80`.

We would read it as:

> Each additional year of education is associated with an estimated 0.80-unit increase in hourly wage in this sample.

The real coefficient is whatever R estimates from the data.

### Why units matter

Never write:

> “Education has an effect of 0.8.”

Write what the number means:

> “One additional year of education is associated with an estimated ___ increase in hourly wage.”

## What about the intercept?

The intercept is the model's predicted wage when:

```text
education = 0
```

That may be mathematically necessary but substantively uninteresting if zero years of education is far from the part of the data we care about.

A coefficient can exist mathematically without being the quantity we want to discuss substantively.

---

# 5. A group comparison is also a regression

Regression is not only for two numerical variables.

We can use a categorical predictor:

```{r}
# Because sex is a factor, lm() automatically represents its categories
# through contrasts relative to a reference category.
model_sex <- lm(
  wages ~ sex,
  data = analysis_data
)

summary(model_sex)
confint(model_sex)
```

Check the levels of `sex`:

```{r}
# The first factor level is normally the default reference category.
levels(analysis_data$sex)
```

R chooses one category as the **reference category**.

The coefficient for the other category represents an estimated difference from that reference group.

This means that:

```{r}
# A two-group comparison written as a t-test.
t.test(wages ~ sex, data = analysis_data)
```

and:

```{r}
# A closely related comparison written as a linear model.
lm(wages ~ sex, data = analysis_data)
```

are asking closely related questions using different presentations of the same basic comparison.

This is one reason regression is so useful: many apparently different statistical questions can be expressed in one modelling framework.

## Change the reference category deliberately

Sometimes the default reference is not the comparison you want.

```{r}
# Inspect the current factor order before changing anything.
levels(analysis_data$sex)
```

You can choose a reference category explicitly with `relevel()`.

```{r}
analysis_data <- analysis_data |>
  mutate(
    # relevel() changes which category is used as the baseline in the model.
    # It does NOT change anybody's observed category.
    sex = relevel(sex, ref = "Female")
  )

# Confirm the new level order.
levels(analysis_data$sex)
```

Refit the model:

```{r}
model_sex <- lm(
  wages ~ sex,
  data = analysis_data
)

# Compare the coefficients with the previous parameterisation.
coef(model_sex)
```

The underlying fitted values do not change merely because we changed the reference category. What changes is how the group difference is written in the coefficients table.

---

# 6. Multiple regression: what changes after adjustment?

Return to our main model:

```{r}
model_1 <- lm(
  wages ~ education,
  data = analysis_data
)
```

This model estimates the overall association between education and wage.

But education is not the only variable associated with wages.

Let us add age.

```{r}
# The + sign means that both education and age enter as predictors.
model_2 <- lm(
  wages ~ education + age,
  data = analysis_data
)

summary(model_2)
```

Compare the education coefficient in the two models.

```{r}
# Simple model: education is the only predictor.
coef(model_1)

# Adjusted model: age is included as an additional predictor.
coef(model_2)
```

The coefficient for education in `model_2` now has a more specific interpretation:

> the estimated difference in hourly wage associated with one additional year of education, **holding age constant**.

Now add the two categorical variables available in the dataset.

```{r}
# R handles the factor predictors sex and language automatically,
# using one reference category for each.
model_3 <- lm(
  wages ~ education + age + sex + language,
  data = analysis_data
)

summary(model_3)
confint(model_3)
```

Again, compare the education coefficient.

```{r}
# ["education"] selects one named coefficient from each coefficient vector.
# This makes the comparison easier than scanning the full model output.
coef(model_1)["education"]
coef(model_2)["education"]
coef(model_3)["education"]
```

This is a useful habit whenever you build models progressively:

> Do not only ask whether a coefficient is “significant”. Ask how its estimated magnitude changes when the model changes.

## Why can a coefficient change?

Suppose a third variable is associated with both:

- the predictor we care about;
- the outcome we care about.

Then part of the simple association may reflect the structure of that third variable.

This is one route to what is often called **confounding**.

But be careful with the language:

> Adding variables to a regression is not automatically the same thing as identifying a causal effect.

Whether a variable should be controlled for is a **substantive and causal question**, not simply a software decision.

---

# 7. What multiple regression does not solve

It would be tempting to conclude:

> “An extra year of education causes wages to increase by the education coefficient.”

Our analysis does not justify that sentence.

Why not?

Because the data are observational and our small teaching dataset does not contain everything that could matter.

Possible issues include:

- omitted variables;
- selection into different levels of education;
- occupational and sector differences;
- work experience;
- measurement error;
- survey design and weights;
- non-linear relationships;
- interactions between variables;
- missing data mechanisms.

Regression answers a precise conditional association question under a model. It does not create a randomized experiment.

## A quick look at residuals

A model also makes assumptions about the pattern left unexplained.

```{r}
# plot() has several diagnostic views for an lm object.
# which = 1 selects "Residuals vs Fitted".
# We look for obvious structure rather than a completely random cloud.
plot(model_3, which = 1)
```

You do not need to become an expert in diagnostics today. The important habit is:

> Never treat `summary(model)` as the end of the analysis.

A model should be checked against the data and against the substantive research question.

---

# 8. Debrief

Today we moved through a sequence that is useful far beyond this particular dataset:

```text
DESCRIBE THE DATA
        ↓
ESTIMATE A QUANTITY
        ↓
QUANTIFY UNCERTAINTY
        ↓
COMPARE GROUPS
        ↓
FIT A SIMPLE MODEL
        ↓
ADD RELEVANT VARIABLES
        ↓
INTERPRET THE COEFFICIENTS
        ↓
STATE WHAT THE DESIGN DOES — AND DOES NOT — SUPPORT
```

The central R commands were:

```text
summarise()
group_by()
t.test()
lm()
summary()
coef()
confint()
```

The central conceptual move was:

> **Do not ask only whether there is a result. Ask what was estimated, how large it is, how uncertain it is, and what assumptions are required to interpret it.**

In Session 3, we will move to a different problem: data in which observations are **not independent** — for example repeated measurements from the same participant, student, object or stimulus — and introduce mixed-effects models.